Classical Simulation of Quantum Error Correction in a Fibonacci Anyon Code

### 2015-04-24 fib2.tex
more minor edits, added a couple of references. apart from the comments in the paper that need to be resolved, for me this is a solid draft at this point.

\begin{abstract}
Classically simulating the dynamics of excitations in two-dimensional quantum systems that 
support anyons is likely intractable in general because such dynamics are sufficient for 
universal quantum computation. However, processes of interest for the study of quantum error correction in anyon systems are typically drawn from a restricted class that displays significant structure over a wide range of system parameters.
We exploit this structure to classically simulate, and thus demonstrate the success of, an error-correction protocol for a quantum memory based on a universal anyon model, the Fibonacci anyons.  We numerically simulate a phenomenological model of the system and noise processes on lattice sizes of up to 
$128\times128$ sites, and find a lower bound on the error-correction threshold of 
approximately $12.5\%$, which is comparable to those previously known for abelian and 
(non-universal) nonabelian anyon models.
\end{abstract}

@@

Topologically ordered quantum systems in two dimensions offer tremendous promise for long-term storage and processing of quantum 
information~\cite{Kitaev2003}. The topological features of such systems are insensitive to local perturbations~\cite{Bravyi2010, Bravyi2011a, Michalakis2013}, and have excitations displaying anyonic statistics~\cite{Wilczek1990}. These anyons can in general be used as quantum memories or to perform universal topological quantum computation~\cite{Freedman2002, Freedman2002b, Nayak????}.

@@

Almost all of the sizable research effort on active quantum error correction for topological 
systems has focused on the case of abelian anyons~\cite{Terhal2014}. These systems are 
well suited to studying quantum error correction because the dynamics of excitations in 
abelian anyon models (at the RG fixed point) can be simulated numerically very efficiently, allowing lattice 
simulations of decoding with over 1 million sites~\cite{Duclos-Cianci2010}. Standard algorithms to efficiently simulate these systems rely on the abelian nature of these particles, and in particular the fact that abelian anyons cannot be used for quantum computation.

@@

Here we consider a full simulation of quantum error correction in a two-dimensional lattice 
system with Fibonacci anyons, a class of nonabelian anyons that are universal for quantum 
computation~\cite{Wang2010b}. Fibonacci anyons are experimentally motivated as the 
expected excitations of the $\nu=\frac{12}{5}$ fractional quantum Hall 
states~\cite{Slingerland2001}, and can also be realized in several spin 
models~\cite{Levin2005, Kapit2013, Palumbo2014} and composite 
heterostructures~\cite{Mong2014}.

@@

The topological features of a physical anyon model are abstractly described by a unitary 
modular tensor category~\cite{Wang2010b}. This object contains the data describing the 
types of anyonic particles, as well as the results of topological operations such as braiding or 
fusion of these anyons. The Fibonacci anyon model consists of only one non-trivial particle 
type, conventionally labelled $\tau$. Denoting the vacuum by $\mathbb{I}$, we can describe 
the possible fusion outcomes of Fibonacci anyons as $\tau\times\tau=\mathbb{I}+\tau$, 
i.e.~two $\tau$ particles can either fuse to vacuum or to a $\tau$ particle.

@@

The fusion outcome of any particular experiment depends on the history of the particles, in 
particular how they have braided around one another. Braiding operations are treated as 
unitary actions on the fusion space. Pair-creation events should be considered isometries 
that enlarge the fusion space, while fusion events correspond to projective measurements 
of combined charge of some number of particles. For details of the description of these 
processes in Fibonacci anyons, see Ref.~\cite{Trebst2008}. 
\stf{check that this ref is a good one...} 
Braiding of Fibonacci anyons and measurements of fusion outcomes is known to be 
sufficient to implement universal quantum computation~\cite{?}.

@@

We consider encoding a qubit of quantum information in the global degeneracy associated 
with the topology of the manifold of our system. We imagine an idealized Hamiltonian for this 
system of the form
\begin{align}
	H=-\sum_{i\in \Lambda}\proj{\mathbb{I}}_i\;,\label{e:hamiltonian}
\end{align}
with $\proj{\mathbb{I}}_i$ the projector to charge $\mathbb{I}$ at site $i$, i.e.~the ground 
space of the model has vacuum (or no anyon) at each site. Given the toroidal boundary 
conditions, this space is two-fold degenerate as noted above. We identify this ground space 
as the codespace of our model. Typical realistic noise processes in this kind of system are 
pair-creation, hopping, exchange etc.~of anyons on neighbouring sites of $\Lambda$. It was 
seen in~\cite{Brell2013} that a simplified noise model consisting of pair-creation events only 
is sufficient to capture the qualitative features of an error-correction simulation, and so we 
will restrict to pair-creation noise processes in this study for convenience. These events act 
on neighbouring pairs of sites, and so may be associated with an edge of the lattice, chosen 
uniformly at random at each time step of the simulation. The simulation time, and thus the 
error-correction threshold, will be measured in terms of average number of noise processes 
per edge, as opposed to an iid noise probability per edge. The latter measure is not 
appropriate for error processes in nonabelian anyon models, but the two coincide in the limit 
of low error rates in those cases where they are comparable (see Ref.~\cite{Brell2013}).

In order to perform a logical error on our code, a noise process must have support on a 
ribbon running around a homologically nontrivial loop of the lattice. These correspond to 
processes in which anyonic charge is transported around the non-trivial loop before 
annihilating to vacuum. More details can be found in Ref.~\cite{?}.

Our error-correction algorithm is based on the hard-decision renormalization group (RG) 
decoders studied in Refs.~\cite{?}. Our decoder proceeds by measuring the occupations of 
anyons at each site of the lattice, forming clusters of nearby anyons and fusing the anyons 
within each cluster, before iterating at increasing length scales. It terminates when the lattice 
is free of anyons (i.e.~we have returned to the codespace) or the clustering length scale 
exceeds the total size of the lattice. \cggb{A few more details? What is our length schedule?} 
\cggb{How does our decoder relate to other decoding schemes used in~\cite{Wootton2013,Brell2013,Hutter2014}?} 
Note that we expect that our qualitative results may be reproduced by most alternative 
families of decoders, such as matching decoders~\cite{?} and soft-decisions 
RG decoders~\cite{?} \cggb{do we?}. The advantage of using a hard-decision RG decoder is 
its simplicity and flexibility, and the fact that its clustering scheme does is compatible with the structure in the noise processes that allows us to classically simulate them.

@@

Below the (bond) percolation threshold for (say) a square lattice, we expect random sets of 
bonds to decompose into separate clusters of average size $O(\log(n))$ and variance 
$O(1)$~\cite{Bazant2000}. \cggb{Do we want to use the word cluster here again or switch?} 
Each noise process in our model is associated with a (randomly distributed) edge, and so 
disconnected clusters correspond to sets of anyons that could not have interacted at any 
point in their history. As such, we can simulate the braiding processes of each cluster 
separately. In other words: the quantum state in the fusion space of all anyons factorizes into 
a tensor product over bond clusters, and we can compute each factor separately. Since each 
cluster has size only $O(\log(n))$, we can typically simulate these dynamics efficiently 
because the resulting fusion space has dimension $O(\mathrm{poly}(n))$. There are rare 
random processes that violate this reasoning, but processes such as these are suppressed 
exponentially in $L$, the lattice size~\cite{Grimmett1989}.

@@

Our concrete simulation proceeds by associating a \emph{tile} with each site of the 
$L\times L$ periodic square lattice. In order to keep track of the observables of the system 
we maintain a set of disjoint curves, representing the history of anyons on the lattice (these 
curves can also be thought of as a top-down view of the fusion tree). 
\stf{this next sentence is currently incorrect.}
The ends of curves 
correspond to leaves of the fusion tree, or anyons. In general, the curves wind around the two 
dimensional tiles in a haphazard way determined by the progress of the simulation. This 
allows a dynamically generated basis for the fusion space that easily allows for tensor 
decomposition into factors corresponding to disjoint clusters. For comparison, previous work 
used a single fixed basis for the fusion space (determined by row-major ordering of 
sites)~\cite{Brell2013}.

>>>>

\begin{abstract}
Classically simulating the dynamics of excitations in two-dimensional quantum systems that 
support anyons is likely intractable in general because such dynamics are sufficient for 
universal quantum computation. However, processes of interest for the study of quantum error correction in anyon systems are typically drawn from a restricted class that displays significant structure over a wide range of system parameters.
We exploit this structure to classically simulate, and thereby demonstrate the success of, an error-correction protocol for a quantum memory based on the universal Fibonacci anyon model.  We numerically simulate a phenomenological model of the system and noise processes on lattice sizes of up to 
$128\times128$ sites, and find a lower bound on the error-correction threshold of 
approximately $12.5\%$, which is comparable to those previously known for abelian and 
(non-universal) nonabelian anyon models.
\end{abstract}

@@

Topologically ordered quantum systems in two dimensions offer tremendous promise for long-term storage and processing of quantum 
information~\cite{Dennis2002} \cggb{a relevant review reference?}. The topological features of such systems are insensitive to local perturbations~\cite{Bravyi2010, Bravyi2011a, Michalakis2013}, and have excitations displaying anyonic statistics~\cite{Wilczek1990}. These anyons can in general be used as quantum memories or to perform universal topological quantum computation~\cite{Kitaev2003, Freedman2002, Nayak2008}.

@@

Almost all of the sizable research effort on active quantum error correction for topological 
systems has focused on the case of abelian anyons~\cite{Terhal2014} \cggb{revise citations here: poulin decoding stuff, what else?}. These systems are 
well suited to studying quantum error correction because the dynamics of excitations in 
abelian anyon models (at the RG fixed point) can be simulated numerically very efficiently, allowing lattice 
simulations of decoding with over 1 million sites~\cite{Duclos-Cianci2010}. Standard algorithms to efficiently simulate these systems rely on the abelian nature of these particles, and in particular the fact that abelian anyons cannot be used for quantum computation.

@@

Here we consider a full simulation of quantum error correction in a two-dimensional lattice 
system with Fibonacci anyons, a class of nonabelian anyons that are universal for quantum 
computation~\cite{Freedman2002, Nayak2008}. Fibonacci anyons are experimentally motivated as the 
expected excitations of the $\nu=\frac{12}{5}$ fractional quantum Hall 
states~\cite{Slingerland2001}, and can also be realized in several spin 
models~\cite{Levin2005, Kapit2013, Palumbo2014} and composite 
heterostructures~\cite{Mong2014}.

@@

The topological features of a physical anyon model are abstractly described by a unitary 
modular tensor category~\cite{Wang2010b}. This object contains the data describing the 
types of anyonic particles, as well as the results of topological operations such as braiding and 
fusion of these anyons. The Fibonacci anyon model consists of only one non-trivial particle 
type, conventionally labelled $\tau$. Denoting the vacuum by $\mathbb{I}$, we can describe 
the possible fusion outcomes of Fibonacci anyons as $\tau\times\tau=\mathbb{I}+\tau$, 
i.e.~two $\tau$ particles can either fuse to vacuum or to a $\tau$ particle.

@@

The fusion outcome of any particular experiment depends on the history of the particles, in 
particular how they have braided around one another. Braiding operations are treated as 
unitary actions on the fusion space. Pair-creation events should be considered isometries 
that enlarge the fusion space, while fusion events correspond to projective measurements 
of combined charge of some number of particles. For details of the description of these 
processes in Fibonacci anyons, see e.g.~Ref.~\cite{Nayak2008} and references therein. 
Braiding of Fibonacci anyons and measurements of fusion outcomes is known to be 
sufficient to implement universal quantum computation~\cite{Freedman2002, Nayak2008}.

@@

We consider encoding a qubit of quantum information in the global degeneracy associated 
with the topology of the manifold of our system. We imagine an idealized Hamiltonian for this 
system of the form
\begin{align}
	H=-\sum_{i\in \Lambda}\proj{\mathbb{I}}_i\;,\label{e:hamiltonian}
\end{align}
with $\proj{\mathbb{I}}_i$ the projector to charge $\mathbb{I}$ at site $i$, i.e.~the ground 
space of the model has vacuum (or no anyon) at each site. Given the toroidal boundary 
conditions, this space is two-fold degenerate as noted above. We identify this ground space 
as the codespace of our model. Typical realistic noise processes in this kind of system are 
pair-creation, hopping, exchange etc.~of anyons on neighbouring sites of $\Lambda$. It was 
seen in~\cite{Brell2013} that a simplified noise model consisting of pair-creation events only 
is sufficient to capture the qualitative features of an error-correction simulation, and so for convenience we 
will restrict to pair-creation noise processes in the numerical results presented in this study. Such pair-creation events act 
on neighbouring pairs of sites, and so may be associated with an edge of the lattice, chosen 
uniformly at random at each time step of the simulation. The simulation time, and thus the 
error-correction threshold, will be measured in terms of average number of noise processes 
per edge, as opposed to an iid noise probability per edge. The latter measure is not 
appropriate for error processes in nonabelian anyon models, but the two coincide in the limit 
of low error rates in those cases where they are comparable (see Ref.~\cite{Brell2013}).

In order to perform a logical error on our code, a noise process must have support on a 
ribbon running around a homologically nontrivial loop of the lattice. These correspond to 
processes in which anyonic charge is transported around the non-trivial loop before 
annihilating to vacuum.

Our error-correction algorithm is based on a hard-decision renormalization group (RG) 
decoder~\cite{Bravyi2011}. The decoder proceeds by measuring the occupations of 
anyons at each site of the lattice, forming clusters of nearby anyons and fusing the anyons 
within each cluster, before iterating at increasing length scales. It terminates when the lattice 
is free of anyons (i.e.~we have returned to the codespace) or the clustering length scale 
exceeds the total size of the lattice. \cggb{A few more details? What is our length schedule?} 
\cggb{How does our decoder relate to other decoding schemes used in~\cite{Wootton2013,Brell2013,Hutter2014}?} 
Note that, as was found in Ref.~\cite{Brell2013}, we expect that our qualitative results may be reproduced by most alternative families of decoders, such as matching decoders~\cite{?} and soft-decisions 
RG decoders~\cite{?} \cggb{do we?}. The advantage of using a hard-decision RG decoder is 
its simplicity and flexibility, and the fact that its clustering scheme is compatible with the structure in the noise processes that allows us to classically simulate them.

@@

Below the (bond) percolation threshold for (say) a 2D square lattice, we expect random sets of 
bonds to decompose into separate clusters of average size $O(\log(n))$ and variance 
$O(1)$~\cite{Bazant2000}. \cggb{Do we want to use the word cluster here again or switch? What was our alternative word again?} 
Each noise process in our model is associated with a (randomly distributed) edge, and so 
disconnected clusters correspond to sets of anyons that could not have interacted at any 
point in their history. As such, we can simulate the braiding processes of each cluster 
separately. In other words: the quantum state in the fusion space of all anyons factorizes into 
a tensor product over bond clusters, and we can compute each factor separately. Since each 
cluster has size only $O(\log(n))$, we can typically simulate these dynamics efficiently 
because the resulting fusion space has dimension $O(\mathrm{poly}(n))$. There are rare 
random processes that violate this reasoning, but processes such as these are suppressed 
exponentially in $L$, the lattice size~\cite{Grimmett1989}.

@@

Our concrete simulation proceeds by associating a \emph{tile} with each site of the 
$L\times L$ periodic square lattice. In order to keep track of the observables of the system 
we maintain a set of disjoint curves, representing the history of anyons on the lattice (these 
curves can also be thought of as a top-down view of the fusion tree). 
\stf{this next sentence is currently incorrect.} \cggb{Why?}
The ends of curves 
correspond to leaves of the fusion tree, or anyons. In general, the curves wind around the two 
dimensional tiles in a haphazard way determined by the progress of the simulation. This 
allows a dynamically generated basis for the fusion space that easily allows for tensor 
decomposition into factors corresponding to disjoint clusters. For comparison, previous work 
used a single fixed basis for the fusion space (determined by row-major ordering of 
sites)~\cite{Brell2013}.
